---
title: "league one dixon-coles 2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##This is verbatim code from: <https://www.statsandsnakeoil.com/2019/01/01/predicting-the-premier-league-with-dixon-coles/> (adapted to use worldfootballR data)

##packages

```{r}
library(tidyverse)
library(worldfootballR)
library(regista) ##https://github.com/Torvaney/regista
library(janitor)

source(here::here("globals.R"))
```

##fix plot size
```{r}
camcorder::gg_record(
  dir = file.path(here::here("camcorder_outputs")),
  device = "png",
  width = 18,
  height = 10,
  dpi = 300
)
```

##font
```{r}
sysfonts::font_add_google(name = "IBM Plex Sans", family = "IBM")

showtext::showtext_auto()
showtext::showtext_opts(dpi = 300)

font <- "IBM"
```


##get data

```{r}
data <-
  worldfootballR::fb_match_results(
    country = "ENG",
    gender = "M",
    season_end_year = 2023,
    tier = "3rd"
  ) |>  
  janitor::clean_names() |>  
  regista::factor_teams(c("home", "away")) |>  
  rename(hgoal = home_goals,
         agoal = away_goals) |>  
  select(10:13)
```

##find unplayed games

```{r}
unplayed_games <- 
  data |>  
  filter(is.na(hgoal) & is.na(agoal))
```

##find played games

```{r}
played_games <- 
  data |>  
  filter(!is.na(hgoal) & !is.na(agoal))
```

##run dixon-coles model

```{r}
model <-
  regista::dixoncoles(
    data = played_games,
    hgoal = hgoal,
    agoal = agoal,
    hteam = home,
    ateam = away
  )
```

##calculate and viz team strength (attack and defence)

```{r}
team_parameters <-
  regista::tidy.dixoncoles(model) |> 
  filter(parameter %in% c("off", "def")) |> 
  mutate(value = exp(value)) |> 
  spread(parameter, value)

team_parameters |> 
  ggplot(aes(x = def, y = off)) +
  geom_point(alpha = 0.5) +
  ggrepel::geom_text_repel(aes(label = team)) +
  theme_minimal() +
  labs(title = "Team strength estimates",
       y = "Attack",
       x = "Defence")
```

##get match probabilities and predicted scorelines for unplayed matches

```{r}
match_probabilities <-
  regista::augment.dixoncoles(model, unplayed_games, type.predict = "outcomes") |> 
  unnest(cols = outcomes) |> 
  spread(outcome, prob)

unplayed_scorelines <-
  regista::augment.dixoncoles(model, unplayed_games, type.predict = "scorelines") |> 
  unnest(cols = scorelines) |>  
  select(-c(agoal, hgoal)) |>  
  rename(agoal = agoal1,
         hgoal = hgoal1)

unplayed_scorelines
```

##bind played and unplayed games with scores

```{r}
played_scorelines <-
  played_games |> 
  select(home, away, hgoal, agoal) |> 
  mutate(prob = 1.0)

scorelines <- bind_rows(
  played_scorelines,
  unplayed_scorelines
)

```

##run single simulation
```{r}
simulate_season <- function(scoreline_probabilities) {
  scoreline_probabilities |> 
    nest(hgoal, agoal, prob, .key = "scorelines") |> 
    mutate(sampled = map(scorelines, ~ sample_n(., 1, weight = prob))) |> 
    select(-scorelines) |> 
    unnest(cols = sampled)
}


single_simulation <- simulate_season(scorelines)
```

##function to calculate table based on played matches
```{r}
calculate_table <- function(games) {
  games_augmented <-
    games |> 
    mutate(
      hpoints = case_when(
        hgoal > agoal  ~ 3,
        hgoal == agoal ~ 1,
        agoal > hgoal  ~ 0
      ),
      apoints = case_when(
        hgoal > agoal  ~ 0,
        hgoal == agoal ~ 1,
        agoal > hgoal  ~ 3
      )
    )

  games_home <-
    games_augmented |> 
    select(
      team   = home,
      gf     = hgoal,
      ga     = agoal,
      points = hpoints
    )

  games_away <-
    games_augmented |> 
    select(
      team   = away,
      gf     = agoal,
      ga     = hgoal,
      points = apoints
    )

  bind_rows(games_home, games_away) |> 
    group_by(team) |> 
    summarise(w  = sum(gf > ga),
              d  = sum(gf == ga),
              l  = sum(gf < ga),
              gf = sum(gf),
              ga = sum(ga),
              gd = gf - ga,
              points = sum(points)) |> 
    arrange(desc(points), desc(gd), desc(gf)) |> 
    mutate(position = row_number())
}
```

##calculate table of played matches
```{r}
tab <- calculate_table(played_games)
```

##calculate table on single simulation
```{r}
p <- calculate_table(single_simulation)
```

##select how many iterations to run
```{r}
n_simulations <- 1000

simulated_tables <-
  rerun(n_simulations, simulate_season(scorelines)) |> 
  map(calculate_table) |> 
  bind_rows(.id = "simulation_id")
```


##save simulated tables
```{r}
write_csv(x = simulated_tables, file = here::here("R", "league_simulation", "league_one_sim.csv"))
```

##current date
```{r}
current_date <- format(Sys.Date(), format = "%d/%m/%Y")
```


##plot
```{r}
(p <- 
  simulated_tables |>
  count(team, position) |>
  group_by(team, position) |>
  summarise(tot = n / n_simulations) |>
  ggplot(aes(
    x = position,
    y = fct_reorder(team,-position),
    fill = tot
  )) +
  geom_tile() +
  scale_x_continuous(breaks = seq(0, 24, 1),
                     expand = c(0.03, 0)) +
  geom_vline(
    xintercept = 2.5,
    linetype = "dotted",
    colour = "red",
    linewidth = 0.7
  ) +
  geom_vline(
    xintercept = 6.5,
    linetype = "dotted",
    colour = "red",
    linewidth = 0.7
  ) +
  geom_vline(
    xintercept = 20.5,
    linetype = "dotted",
    colour = "red",
    linewidth = 0.7
  ) +
  scale_fill_continuous(low = "white", high = "#72aeb6") +
  geom_text(aes(
    label = paste0(tot * 100, "%"),
    size = 2,
    family = font
  )) +
  labs(
    title = "League One Projections - 2022/2023",
    subtitle = glue::glue("Simulation up to: {current_date}"),
    y = "",
    x = "Position",
    caption = glue::glue("Viz: @AnalyticsOxford\nData: Fotmob.com via WorldfootballR\nFollowing: https://www.statsandsnakeoil.com")
  ) +
  theme(
    plot.title = element_text(
      family = font,
      size = 30,
      face = "bold"
    ),
    panel.grid.major = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.text.y = ggtext::element_markdown(size = 14, family = font),
    axis.ticks = element_blank(),
    plot.subtitle = ggtext::element_markdown(size = 16),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14, family = font),
    plot.title.position = "plot",
    plot.caption = element_text(size = 12)
  ))

```

##save plot
```{r}
ggsave(
  p,
  filename = here::here("R",
                        "league_simulation",
                        "league_one_sim_plot.jpg"),
  dpi = 300,
  height = 10,
  width = 18
)
```

##load plot and logo
```{r}
plot <-
  here::here("R",
             "league_simulation",
             "league_one_sim_plot.jpg")

logo <- here::here("src", "efl.png")
```

##position logo on plot
```{r}
(
  all_plot <- add_logo(
    plot_path = plot,
    logo_path = logo,
    logo_position = "top right",
    logo_scale = 10
  )
)
```

##save complete image
```{r}
magick::image_write(all_plot, path = here::here(
  "R",
  "league_simulation",
  glue::glue("league_one_sim_plot_full.jpg")
))
```

